### 2019-05-06 paper-four/paper-four.tex
paper four first post

\usepackage{polyglossia}
\setdefaultlanguage{english}
\setotherlanguage{sanskrit}
\usepackage{fontspec}
\newfontfamily\devanagarifont[Scale=MatchUppercase]{Kohinoor Devanagari}

\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{float}
\usepackage[bottom]{footmisc}
\usepackage{enumitem}
\usepackage{hyperref}

@@

\begin{titlepage}
   \vspace*{\stretch{1.0}}
   \begin{center}
     \Huge\textbf{Vipassana for Hackers}\\
     \vspace{5cm}
     \large\textit{Steven Deobald}\\
     \large\textit{Version 1.6}\\
     \large\textit\today\\
     \vspace{5cm}
     \large\textit{Contributions by: Preethi Govindarajan, Mat Kelly, Liam Monahan}\\
   \end{center}
   \vspace*{\stretch{2.0}}
\end{titlepage}

@@

Should you bother reading this paper? Not everyone associates themselves with the label ``Hacker''. It comes with a few well-known negative connotations so my first task is to quickly identify what is meant by ``Hacker'', since this paper is for these people.

My first exposure to the term ``hacking,'' outside of the late-80's imagery which became fodder for the 1995 film, was through my cousin Jennie. She is an Ecologist, a self-described ``Mud Scientist'', who most people would not associate with the term ``Hacker''. She visited MIT during her PhD to find the term extended to ``roof and tunnel hacking,'' an idea and practice which delighted her to no end. MIT or not, Jennie is a Hacker. Every Science is about exploration and discovery, and these are the concepts at the root of ``hacking.''

Carpentry, painting, gardening, the entire field of Genetics, baking bread, Philosophy, electronics, fashion, Psychology, health, writing, and bicycles. Almost any discipline can fall under the banner of ``hacking'' I'm using for this paper, since almost any discipline can (but does not necessarily) include the potential for two essential aspects of the Hacker Mentality: Rationality and Curiosity.

This is all that is required.

If you have ever enjoyed taking apart a computer or car, experimenting with bread recipes, researching and writing, pulling apart the semantics of your work, building something, rebuilding something, exploring a cave or abandoned construction site... I wrote this paper for you. If you enjoy any exploratory vocation or hobby you will almost certainly find Vipassana utterly fascinating. The reason is this: As acutely as a human being can disassemble a stereo, examine the slide of a microscope, or explore a cave, human beings have the capacity to disassemble, examine, and explore ourselves.

What may be even more fascinating, for the hacker, is the fact that by practicing Vipassana one can seemingly increase the capacity for rational thought and behaviour.

I have written this paper because, thus far, I have found the written and recorded material on the subject of Vipassana to fall into three broad categories:

Well-meaning Neuroscience papers which completely miss the mechanics of nervous-system-based meditation techniques and their core outcomes.
Descriptions of a ``silent meditation retreat,'' usually given by someone who has not attended a course. \cite{thenewsroom}
Descriptions of Vipassana from an individual and subjective perspective, usually given by someone who has attended one course. (``Oh my god! It was sooooo intense!'')

By using phrases like ``disassemble, examine, and explore ourselves'' I'm at the risk of sounding spiritual. This is the opposite of my intention. My intention is to pique curiosity in fellow hackers by explaining the biological mechanics of Vipassana, to the best of my ability and as I understand them. To curb the impression that this paper falls into the category of New Age Philosophy, I would like to take a moment to address what this paper is not.

\pagebreak

\begin{center}
  \Huge{What This Paper is Not}
\end{center}

This not an instruction manual. You will not learn Vipassana from reading this paper.

This is not a white paper. I am not a scientist. This is not science.

This is not an expert's treatise on the subject of nervous-system-based meditations, nor even an expert Vipassana meditator's description of the technique. I am not an expert. I have only attended four 10-day courses thus far. Each has been substantially different from the others.

This paper will endeavour to avoid falling into the trap of ``Buddhist Guys'' (a mildly derogatory term coined by a serious meditator friend of mine). You might be familiar with phrases commonly attributed to this archetype such as ``When the student is ready, the teacher will appear.'' Mysticism, mystery, and mystique are to be avoided. If I fall into this trap, email me: \href{mailto:steven@nilenso.com}{steven@nilenso.com}

This is not a discussion on ``spirituality'', for any definition of the term.

This is not a dissection of the administrative details of the course, though I will provide a brief overview of the course schedule.

This is not a description of past experiences, though I will use a few to highlight core effects of the technique.

This is not a comparison with other meditation techniques, though I will address other techniques for context. It is worth noting that there are a few meditation techniques labelled ``Vipassana'' and this paper deals specifically with Vipassana as taught by S.N. Goenka.

This is not an examination of Buddhist Philosophy and will exclude commentary or mention of the following concepts:

\begin{itemize}
  \item Siddhartha Gautama (The Buddha)
  \item Enlightenment
  \item Religion / Buddhism
  \item Prophecy / Prediction of the future
  \item Telepathy / ESP
  \item Reincarnation
  \item ``Energies''
  \item Kalapas
  \item Metta / Morality
  \item Sankharas / ``Impurity''
\end{itemize}

These last two bullet points are subject to one exception each.

\pagebreak

\begin{center}
  \LARGE{Morality's Exception: Lying}
\end{center}

A 10-day Vipassana course requires that all students take the standard \textit{5 Precepts} \cite{fiveprecepts}, as per Theravada. I do not recommend getting drunk, raping and murdering your fellow students, and stealing their belongings. However, the four precepts which forbid those behaviours are quite clearly external, making them components of course administration and therefore not applicable to this document.

Lying, however, is perhaps least interesting in its external form. As an educated adult, it is difficult to say anything at all without lying to some small degree. As one begins to systematically observe the nervous system, which is the basic mechanic of Vipassana, one finds that the brain and nervous system are acutely aware of even the tiniest untruth. If I tell myself I am feeling something which I am not (``my body is exploding into a billion tiny particles! my muscles are raging rivers of lava inside my body!'') the process of empirical and objective observation ends, and the process stops.

Therefore, the exception given to Lying (or the precept governing Lying) is because it actually prevents the technique from working, not because I have some particular crusade against lying.

\vspace{1cm}

\begin{center}
  \LARGE{Habit's Exception: Saṅkhāra}
\end{center}

The notion represented by the Pali word \textit{Saṅkhāra} can be split into two camps, in terms of its definition and use. The first, the notion that the body and mind contain ``impurities'' and that the objective of meditation is to rid the body-mind of such impurities, will be excluded from all discussion here. The second, more common use of the term ``Sankhara'' refers to ``conditioned phenomena'' or ``psychological dispositions''. The equivalent Hindi/Sanskrit word, \foreignlanguage{sanskrit}{संस्कार} (sanskaar) roughly translates in this context to ``mental impression''. This second definition will apply to this paper.

Here, we might choose a more specific term from Psychology: mental habits. Saṅkhāras, as far as this document is concerned, can be thought of as the ingrained habits of the brain and nervous system. Some concrete examples include liking (or disliking) the taste of olives, my body's ability to type at this QWERTY keyboard without consciously considering each keystroke, and the physical and emotional response I have to news of violence in my city of residence.

\pagebreak

\begin{center}
  \Huge{A Finger Pointing at the Moon}
\end{center}

The fundamental difficulty in any description of Vipassana is this: It is impossible to convey a description of Vipassana in terms known to someone who has not experienced it. During a Vipassana course this paradox is addressed using two metaphors: First, a story of a sighted brother explaining the concept of ``white'' to his blind brother. \cite{experiencethetruth} Second, the classic Indian story of blind men each describing an elephant and each producing a partial truth.

The first metaphor is a better framework for explaining the difficulty of writing about nervous system meditations. It is fundamentally impossible to express the experience of observing ``whiteness'' to an individual who is blind from birth. This person requires sight to understand. Similarly, it is fundamentally impossible to describe the effects of a nervous system meditation which alters perceptions and the interactions of the nervous system itself, since the nervous system (including the brain) ultimately form our entire experience.

The hacker community has been on the verge of this exploration for quite some time. The easiest and most recent examples of this are no doubt the works of Douglas Hofstadter. \textit{Gödel, Escher, Bach: An Eternal Golden Braid} captured the imaginations of the hacker community and grew from esoteric pub conversation piece to dinner party staple, around the same time as the term ``blogging'' evolved from portmanteau to an accessible part-time career for stay-at-home parents. The term ``strange loop'', coined in this book, has since become a mild obsession of philosophical and artistic computer scientists, linguists, and mathematicians, even inspiring the namesake software conference held in St. Louis. Hofstadter's more cohesive \textit{I Am a Strange Loop} tackles the problem of consciousness more succinctly. However, other than directing the reader to Zen\footnote{Zen is relevant because the practice of Zazen (Zen meditation) is a nervous-system-based meditation similar to Vipassana. The reason I do not discuss Zazen in detail or compare it to Vipassana in this paper is because Zazen, in my experience, is not systematic. The subject of inspection — the nervous system — remains, but Zen teachers tend to avoid explicit instruction, leaving the student to explore on her own. The outcomes, therefore, appear similar but watered down.} at multiple points, these books do not invite the reader to explore these concepts directly. There is no hacking. A thorough read of GEB or IAaSL is akin to reading all four volumes of Knuth's \textit{The Art of Computer Programming} \cite{taocp} without having ever touched a computer or to reading \textit{The Joy of Cooking} \cite{joyofcooking} cover to cover without having ever set foot inside a kitchen.

\pagebreak

\begin{center}
  \Huge{Computing Sidebar: LISP}
\end{center}

I will attempt to frame the self-exploration of nervous system meditations like Vipassana using one tool familiar to the Computer Science discipline: LISP. If you are neither a Computer Scientist nor a Linguist and have no interest in computer languages, feel free to skip to the next section.

Eric Raymond described the exercise of LISP mastery to be ``profound enlightenment'' \cite{hackerhowto} and while we can probably agree to ignore the hyperbole, there is something to be said about a language which begins as a calculus and nearly manages to escape formal grammar entirely. LISP has some very interesting properties:

\begin{itemize}
\item \textbf{Homoiconicity:} the language is represented in itself and code written in the language is understood by the language
\item \textbf{Metaprogramming:} the language can program itself and understand itself
\item \textbf{Self-mutation and self-replication:} a LISP program can inherently alter its own internal behaviour and easily write entirely new programs
\item \textbf{Recursion and trampolining:} mathematical functional recursion and mutual recursion between two functions are natural operations in LISP
\end{itemize}

Combined, these properties tend to blur the lines of compile time and run time, of code and data, of execution and interpretation. Decades ago, thanks to these ``strange loop'' properties of LISP, it was thought to herald the dawn of the Age of Artificial Intelligence. While that did not happen it left us with a fascinating building material: LISP appears to have no upper-bound. It is ``turtles all the way up.'' LISP is not ``turtles all the way down'', as some would describe it. Even if written on a LISP Machine (a computer which executes LISP directly rather than transforming LISP into machine-level instructions), there are still paths of wires and electrons which need to be dealt with. These are not written in LISP.

Vipassana would appear to be ``turtles all the way down'' in precisely this sense. Vipassana is the process of digging through your consciousness using consciousness itself. At no point do you reach the machine beneath the machine: ``Aha! I found it! The Consciousness Machine. Now I can do away with the meditation and just read the manufacturer's open specification and I'm sorted.'' \textit{The machine at the bottom of your consciousness is a strange loop}. Having a look at this strange loop is a lot more fun than reading about it.

\pagebreak

\begin{center}
  \Huge{Course Structure}
\end{center}

For those unfamiliar, I will briefly overview the rules of the course. During the course you are not permitted books, phones, talking, looking people in the eye, touching, making sounds, snacks, music, or writing implements.

Put more succinctly, your activities will include: Eat, sleep, poop, bath, meditate.

That's it.

Meditation will consume approximately ten to eleven hours per day. All courses the world over are free and volunteer-run. Material goods are provided for by donations from previous students.

\pagebreak

\begin{center}
  \Huge{Suspending Disbelief}
\end{center}

Before explaining the mechanics and outcomes of Vipassana, I must request that you suspend disbelief about three items for the duration of the paper.

\vspace{1cm}
\begin{center}
  \LARGE{1. Meditation is a thing}
\end{center}

For the purposes of this document, understand ``meditation'' to mean a single point of awareness with unwavering focus. ``Unwavering focus'' means that we can hold sustained attention on this single point (whatever it may be) without getting distracted by thoughts of our taxes, what's for lunch, or that really persistent itch on our ear.

What I mean by this is that mental exercises exist which can take consciousness beyond ``flow'', ``calm'', or ``elated'' states of mind caused by activities one might describe as ``meditative'' rather than ``meditation''. Biochemical responses to hallucinogens fall on the same side of the spectrum, though they may appear momentarily insightful.

I have drawn my personal spectrum for sample activities in Figure~\ref{fig:meditation-spectrum-1}.

\begin{figure}[H]
  \centering
  \includegraphics[width=11cm]{images/meditation-spectrum-1.jpeg}
  \caption{Sample meditation/meditative spectrum.}
  \label{fig:meditation-spectrum-1}
\end{figure}

For the purposes of this paper, we will keep all notions of ``mindfulness'' in the ``meditative'' camp, on the right. Mindfulness is not meditation. This paper will be discussing only two practices from this list: Anapana and Vipassana. An illustrated definition of ``awareness'' and ``attention'' follows, in the section entitled ``On Attention''.

\begin{figure}[H]
  \centering
  \includegraphics[width=11cm]{images/meditation-spectrum-3.jpeg}
  \caption{Sample meditation spectrum.}
  \label{fig:meditation-spectrum-3}
\end{figure}

\vspace{1cm}
\begin{center}
  \LARGE{2. Your brain and nervous system have}\\
  \LARGE{a tight feedback loop}
\end{center}

When we say ``tight feedback loop'', this means the brain and nervous system are tied together in a feedback loop faster and tighter than you could ever imagine without experiencing it directly.

We all know the classic examples of connectivity between the brain and nervous system: A child touches a hot stove and retracts her hand. A pubescent boy sees a boy or girl he is attracted to and feels ``butterflies in his stomach'' so intense he vomits. Despite healthy diet and exercise, a high-strung individual suffers from stomach ulcers.

These examples all exist on a gross level, coarse-grained enough that we can identify them externally and even give them names like ``butterflies''. However, your nervous system is interacting with your brain at every conceivable level. No matter how small, you do not have a thought or emotion emerge in your consciousness (or unconscious) without a physical sensation accompanying it. You do not feel physical sensation, consciously or subconsciously, without an accompanying thought or emotion. This interaction is continuous.

Unless you have tried a meditation system with the nervous system as its meditation object, you have almost undoubtedly not experienced this.

\vspace{1cm}
\begin{center}
  \LARGE{3. Dissolution of the body is a thing}
\end{center}

This will be the most difficult of the three. ``Dissolution of the body,'' though strange-sounding at first, does not fall into the categories where we would normally place magic, levitation, or ESP. The term simply refers to our capacity to see through our entire body.

As we investigate this concept, it seems increasingly reasonable in conjunction with points \#1 and \#2. If we can narrow our attention down to anything we can perceive and what we choose to perceive with our brain and nervous system in tandem is a narrow part of the body, surely if we sustain attention on that part for long enough we will start to see it clearly.

Normally, our attention on our body is almost entirely external: my hands on this keyboard, the itch on my ear, my butt in this chair. I don't tend to spend a lot of energy focused on the sensations provided to my brain by my gallbladder or my stomach unless I have a painful gallstone or I am experiencing acute hunger. Physical sensations provided to our awareness by the nervous system for the internals of the body are largely hidden from us — but they can be unearthed and explored.

I should note here that in four courses I have not felt the precise shape of my gallbladder or stomach at any point. I have ``seen'' the shape of my bones in my fingers, shoulders, chest, and pelvis. I have witnessed the rush of blood squooshing through my cardiovascular system (near the surface) with terrifying clarity. I have enough anecdotal evidence to say that it appears this process does work and can be extended to our deepest internal organs and the spinal cord.

To help visualize what I mean, a sample spectrum of sensations from the gross or obvious to the curious and subtle might look like the figure on the following page.

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{images/sensation-spectrum.jpeg}
  \caption{Sample sensation spectrum.}
  \label{fig:sensation-spectrum}
\end{figure}

As the field of sensation progresses to the left, the sensations felt become increasingly difficult to describe. ``Surreal hot / cold'' refers to sensations of temperature we wouldn’t normally consider possible: our muscles either on fire or freezing solid. ``Spatial distortion'' refers to complete alterations of spacial presence — the partial or complete replacement of one’s normal sense of proprioception. This may include reorientation in space (``I was upside-down!'') or the feeling that one is gigantic or tiny or that any one body part has one of these or other distorted qualities.

``Throbbing'' refers to thick, heavy, slower vibrations; these are often felt rhythmically and this rhythm accelerates until the sensation is described as a ``vibration''. Beyond vibration is the ``invisible'', where a particular body part is no longer seen and a deeper part of the body is experienced directly. The first such example in most cases tends to be the skin, since the external surface of the body is the first part of the nervous system the course instructions ask the meditator to focus on. Beyond this the meditator may find muscle tissue, bones, the cardiovascular system, or something else entirely. Usually by the point in the course when a meditator has dissolved her own skin to her attention the sensations become extremely difficult to identify and label so ``something else entirely'' is likely to be the most common case.

The sample spectrum above is missing many, many sensations. I also feel it necessary to call out that the spectrum is not serial; although I have felt ``vibrations'' and seen deeper than skin level, I have certainly not experienced all the sensations possible in the spectrum. However, the diagram hopefully provides a reference point for the type of readily-accessible sensations which Vipassana begins with (on the right) and the types of sensations (or absence of sensation) which will convince a person that complete dissolution of the body not only possible but an entirely natural process.

\pagebreak

\begin{center}
  \Huge{Virtual Reality}\\
  \Huge{and}\\
  \Huge{The Matrix Paradox}
\end{center}

The following section will require some visual assistance, for which I will use the same slides I used in the talk which corresponds to this paper. An unmodified slide is seen here in Figure~\ref{fig:michael-abrash-senses}.

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{images/michael-abrash-senses.png}
  \caption{Traditional 5-sense brain mapping.}
  \label{fig:michael-abrash-senses}
\end{figure}

This image is taken from a talk by Michael Abrash, Chief Scientist of the Oculus VR team, at a talk he gave at the Facebook F8 Summit in 2015. \cite{abrashvr} I'm using this slide because it's nicely laid out and because it is incomplete.

Abrash discusses the nature of the ``Matrix Paradox'' (more commonly known as the \textit{Simulation Hypothesis} \cite{simulationhypothesis}) as simply the right neurons firing to simulate any particular environment. ``How do we know we are not in the matrix?'' The question is valid, but the diagram is missing a great deal of sensory input. Figure~\ref{fig:michael-abrash-senses-extended} is a slightly better model of reality.

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{images/michael-abrash-senses-extended-attached.jpeg}
  \caption{A more complete, 7-sense brain mapping.}
  \label{fig:michael-abrash-senses-extended}
\end{figure}

Though a classic sense diagram such as Mr. Abrash's might describe proprioception \cite{ramachandranbrainfunction, proprioception}, nocioception \cite{nocioception}, and hunger \cite{fmrihunger} as physical ``feeling'', they are definitively not a subset of ``touch'', as they correspond to entirely different activity in the brain. Touch is strictly an external sense dealing with skin. These are internal senses.

A brain diagram describing The Matrix Paradox is complete only when we map all senses to the brain. I have not, since I haven't named them all. In fact, neurobiology does not yet have names for all sensory input, making such an exercise exceedingly difficult. Proprioception is clearly associated with the cerebellum and hunger is clearly associated with the hypothalamus. Nocioception, however, is a bit more nuanced with regard to the brain's reception of thermal, mechanical, and chemical information — and whether or not the activity is inbound, outbound, conscious, or unconscious. However, assuming all nervous system activity does have corresponding brain activity, one could theoretically map out the Matrix Paradox on the brain as Abrash has started. It seems likely such a diagram would highlight all parts of the entire brain.

On that note, I feel I should introduce an absolutely fantastic RadioLab episode (\textit{Where Am I?} \cite{whereami}) which addresses this topic of ``other senses'' from the perspectives brought by modern Neuroscience. On the topic of lacking proprioception:

\begin{quote}
  \textbf{We have words like `deaf' and `blind'. We do not have words for being deaf or blind to one's own body.}
\end{quote}

In the Matrix, perhaps this is either a bug or a feature flag.

Now that we have our diagram for visualizing the use of our senses, we can address them in terms of various altered mental states. Before we get there, however, we will try to come up with a meaningful visual definition of ``attention'' or ``awareness'' with respect to our senses.

\pagebreak

\begin{center}
  \Huge{On Attention}
\end{center}

The funny thing about Vipassana is that you are doing it already... you're just terrible at it. Vipassana is the inspection of one's own nervous system with narrowly-focused attention but the precision of this focus is not prescribed by the instruction given for the technique. ``As narrow as possible'', perhaps, but in day to day life a normal person cannot possibly narrow her attention to the size of a fingertip. Your attention is, in part, already on your nervous system, with some degree of focus. To the extent that it is, you are doing Vipassana.

If your attention is not directed at your nervous system, where is it? Let's look at three environmental examples which can help us highlight our senses and our awareness within them.

\pagebreak

\begin{center}
  \LARGE{Example 1:}\\
  \LARGE{A bustling restaurant}
\end{center}

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{images/attention1-restaurant.jpg}
  \caption{A bustling restaurant.}
  \label{fig:bustling-restaurant}
\end{figure}

...and a corresponding sense map something like this:

\begin{figure}[h]
  \centering
  \includegraphics[width=10cm]{images/ma-noisy-restaurant.png}
  \caption{Sense map of a bustling restaurant.}
  \label{fig:sense-map-bustling-restaurant}
\end{figure}

Assuming we've just placed food in our mouths, our sense of taste is going off like crazy, we're hearing a lot of scattered sound around us, and we're also sufficiently distracted by most of our other senses. Note that we are probably paying very little attention to the internal senses.

\pagebreak

\begin{center}
  \LARGE{Example 2:}\\
  \LARGE{A sensory deprivation tank}
\end{center}

\begin{figure}[h!]
  \centering
  \begin{subfigure}[b]{0.48\linewidth}
    \includegraphics[width=\linewidth]{images/attention2-tank.jpg}
  \end{subfigure}
  \begin{subfigure}[b]{0.48\linewidth}
    \includegraphics[width=\linewidth]{images/ma-sensory-deprivation-tank.png}
  \end{subfigure}
  \caption{A sensory deprivation tank and corresponding sense map.}
  \label{fig:sensory-deprivation-tank}
\end{figure}

Here, the classic 5-sense model fails us, since sensory deprivation only deprives us of these 5 senses. When Neo is first exposed to the reality of The Matrix, the pods which imprison the humans loosely resemble such sensory deprivation tanks, which is reason enough to believe that we are not in that matrix. As anyone who has ever entered a sensory deprivation tank knows, it is not like being in a coma. The mind is still very active and it may actually help in causing someone to pay attention to their internal senses.

\pagebreak

\begin{center}
  \LARGE{Example 3:}\\
  \LARGE{An anechoic chamber}
\end{center}

\begin{figure}[h!]
  \centering
  \begin{subfigure}[b]{0.48\linewidth}
    \includegraphics[width=\linewidth]{images/attention3-chamber.png}
  \end{subfigure}
  \begin{subfigure}[b]{0.48\linewidth}
    \includegraphics[width=\linewidth]{images/ma-anechoic-chamber.png}
  \end{subfigure}
  \caption{An anechoic chamber and corresponding sense map.}
  \label{fig:anechoic-chamber}
\end{figure}

The map for an anechoic chamber may seem surprising. However, just because the room you are in has a negative decibel rating does not mean you cannot hear. In fact, in this environment, your sense of hearing is likely to be amplified --- and turned inward. You will hear your heart beating, your blood pumping, your digestive system sloshing around. The better the anechoic chamber at swallowing sound, the more intense this experience is going to be.

For this reason, the above anechoic chamber at Orfield Laboratories \cite{orfield} has an unofficial record of 45 minutes for the longest period anyone has sat inside it. As it turns out, hearing one's own bodily functions at deafening internal volumes is likely to cause someone to go a bit crazy, hence the highlighted bar for thought activity.

The above three examples serve as templates with external input guiding one's understanding of what is happening neurologically.

\pagebreak

\begin{center}
  \Huge{Anapana Meditation:}\\
  \Huge{Practicing Focus}
\end{center}

The first 3.5 days of a 10-day Vipassana course are spent practicing ``Anapana'' (breath) meditation. This period is very important as it is time spent practicing acute, one-pointed attention. If we visualize our attention as any skin (internal or external; lips, nose, nostrils, etc.) covered by the pink triangle in Figure~\ref{fig:burmese-anapana-triangle}, we will get an idea of exactly where one's attention is directed at the beginning of Anapana practice.

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{images/burmese-anapana-triangle.png}
  \caption{Large, triangular Anapana meditation target.}
  \label{fig:burmese-anapana-triangle}
\end{figure}

If you have ever practiced a single-pointed meditation (particularly breath meditations) before, you know all too well how easily your mind can be distracted. Taxes, Mom's birthday, an upcoming meeting, a fun math problem, what's for lunch, the itch on my ear... anything can become a distraction because everything which is not ``my breath'' is by definition a distraction. And we are not accustomed to spending long stretches of time with unwavering, singular focus. We are especially unaccustomed to doing so with all of our attention focused on our breath.

As you practice Anapana for over 30 hours, you will find your attention wandering less and less, while (paradoxically) the distractions become more and more interesting. The itch on your ear becomes an itch on your nose which seems to move with the breath. Your brain will spend less time worrying about taxes and more time trying to solve interesting math problems (if that's your thing).

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{images/face-attention.jpg}
  \caption{Wandering attention.}
  \label{fig:wandering-attention}
\end{figure}

On the first day, your attention might look something like Figure~\ref{fig:wandering-attention}, where the wavy pink cloud represents any attention spent on any part of the body other than the nostrils or spent ``outside'' the body: on thoughts and emotions. By the fourth day, your attention will often take a shape something like Figure~\ref{fig:narrow-attention}. Over the course of the first three days, the instructions narrow down one's attention from the nose-and-lips triangle from Figure~\ref{fig:burmese-anapana-triangle} to a finger-tip-sized area wherever the breath can be felt beneath the nostrils.

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{images/face-attention-narrow.jpg}
  \caption{Narrow attention.}
  \label{fig:narrow-attention}
\end{figure}

If we revisit the notion that ``meditation is a thing'', we can see a specific example in Anapana with the help of the sense map in Figure~\ref{fig:anapana-sense-map}. Where mindfulness and violin, albeit useful physical and mental practices, are not meditation, we can see from the sense map that Anapana \textit{is} meditation. It is the practice of narrowing one's attention, repeatedly, toward a specific sense (in this case, touch) at a specific time (now) in a specific place (just beneath the nostrils). As our skill in Anapana improves, we come nearer and nearer to the focus-oriented meditation ideal of an unwavering single point of awareness.

\begin{figure}[H]
  \centering
  \includegraphics[width=10cm]{images/ma-anapana.png}
  \caption{Anapana sense map.}
  \label{fig:anapana-sense-map}
\end{figure}

In Figure~\ref{fig:anapana-sense-map}, we can see that in reality our attention is not perfectly tuned and narrowed to a tiny spot under our nose. We are still likely to hear sounds around us on the third day of the course, especially if the sound is monkeys clambering over the roof of the meditation hall. By the third day of the course, however, one's attention to sound is likely to be greatly reduced.

For the rest of this paper, make note of the fact that \textit{distraction} is at this stage, and will forever remain, the most important aspect of meditation; it is everything meditation is not, and therefore a significant consideration in all the techniques which follow. Meditation is essentially comprised of only two components: (a) the meditation object and (b) distractions. The activity of meditation is the repeatedly directing of one's attention from the latter to the former.

\pagebreak

\begin{center}
  \LARGE{Anapana: The Bridge}
\end{center}

At the end of the first day of a 10-day Vipassana course, the breath is described as a bridge ``from the known to the unknown, because respiration is one function of the body that can be either conscious or unconscious, intentional or automatic.'' During the 3-day Anapana session of a 10-day course, students are instructed to \textit{consciously} direct attention toward their breath -- and even gently control the breath for brief periods to maintain attention when the mind won't stop wandering -- until awareness of breath, without force, lasts for extended periods. Given that the breath is the one function of the body which operates both consciously and unconsciously, it makes sense that it acts as the starting point for many meditation practices, including Vipassana. The intention of Vipassana is to train the mind out of habits and tendencies which have embedded themselves into the brain and nervous system at an unconscious level. To do this, one must first learn to access one's unconscious. Through the breath, Anapana meditation fulfills this purpose.

Vipassana requires a 10-day course to learn. Anapana does not. Before his death, S.N. Goenka released both 70-minute audio instructions as MP3s for the Anapana meditation technique.\cite{anapana} The 70-minute instructions can be found at \url{http://www.vipassana.co/What-is-Anapana} under ``Anapana Meditation for All'' and come as four 15-minute sessions plus a 10-minute ``daily practice'' session.

\begin{figure}[H]
  \centering
  \includegraphics[width=8cm]{images/bridge.png}
  \label{fig:bridge}
\end{figure}

\pagebreak

\begin{center}
  \Huge{Vipassana Meditation:}\\
  \Huge{Mutually-Recursive}\\
  \Huge{Consciousness I/O}
\end{center}

Oddly enough, Vipassana mostly seems to be about remaining calm. If you're thinking ``you're sitting there, silent, for 10 hours a day... what is there to get upset about?'' you are in for a surprise. But first, a quick overview of the early mechanics of Vipassana. (See Figure~\ref{fig:vipassana-mechanics}.)
Attention is moved from the nostril area to the top of the head, then systematically moved across every body part, down to the toes. This process then repeats.

This visualization and description is to act as an \textit{example only}: this is not the only way to begin Vipassana meditation. Though Vipassana instruction is systematic, it is far from mechanical or prescriptive. Even in the initial instructions, alternative approaches are offered including alternate orders of observation, alternate areas of observation, and alternate durations of observation. The key aspect of any approach to Vipassana is that the \textbf{entire body} is repeatedly observed within each meditation period.

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{images/burmese-vipassana-mechanics.png}
  \caption{Simplified example of initial Vipassana mechanics.}
  \label{fig:vipassana-mechanics}
\end{figure}

I will take this opportunity to echo a caveat from the beginning of this paper: This section is in no way a form of instruction. Do not attempt full-body Vipassana meditation outside of a proper 10-day course. Practicing Vipassana is likely the most difficult activity you will ever attempt, should you choose to. To try so alone, you will have no teacher for guidance and the experience is bound to be frustrating.

Consider this warning to also extend to any literature which \textbf{does} attempt to convey Vipassana meditation instruction. Specifically, Glickman's \textit{Beyond the Breath: Extraordinary Mindfulness Through Whole-Body Vipassana Meditation} \cite{glickman} includes a facsimile of instruction provided on a 10-day introductory Vipassana course. Instruction on a course includes 11+ hours of video instruction, at least twice as many hours of audio instruction, and multiple interviews per day with the teacher. Trying to replace this with a book is both inadequate and dangerous.

More complex variations are added to these mechanics as the bodily sensation experienced through this probing of the nervous system becomes increasingly subtle. In this order:

\begin{enumerate}
  \item Add toes-to-head to head-to-toes
  \item Dual/symmetrical observation (both arms, both legs)
  \item Observation of a ``ring'' around the entire body
  \item Observation of as much of the body as possible (while moving attention up and down)
  \item Penetrating attention of muscles, tissues, organs
  \item Penetrating attention of the spinal cord
  \item Randomized attention in fingertip-sized dots throughout the body, inside and out
  \item Sweeping attention through the entire body, up and down, in a single breath
\end{enumerate}

Through this process, attention will move from the sense of touch to all the internal senses and the sense map will look more like Figure~\ref{fig:vipassana-sense-map-1}.

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{images/ma-vipassana1.png}
  \caption{Vipassana sense map.}
  \label{fig:vipassana-sense-map-1}
\end{figure}

I should note that I have never successfully penetrated my muscles or inner tissues much. The average new meditator is likely to feel a great many fascinating sensations on her fingertips, palms, arms, toes, and legs. The torso, since it is surrounded in longer nerve fibres, will take longer to observe in the same way. \textit{[citation needed]}

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{images/burmese-vipassana-pain.png}
  \caption{Common pain for Westerners in meditative postures.}
  \label{fig:burmese-vipassana-pain}
\end{figure}

As a Canadian who spent little (if any) of my childhood sitting cross-legged on the floor, Figure~\ref{fig:burmese-vipassana-pain} roughly summarizes most of my early meditation experiences, where the purple dots represent intense pain and the pink dot continues to represent my awareness. At the peak of my meditation practice toward the end of a course, I would find that if my attention were somewhere away (say, my neck) from the painful area of my body (say, my knees), I may not feel any pain at all. As I moved my attention around, the pain in my knees would come back as I approached them, but every time I could sustain focus on another part of my body, the pain would appear to decrease with every return visit to my knees. If I can remain calm and focused while away from my knees, my knees' pain seems to take care of itself. If I can remain calm and focused while my attention moves across my knees, the situation continues to improve.

If an individual cannot feel pain in one's knees despite the fact that the pain is ``real'', she has effectively eliminated (some) sensory input. This is possible for all sensory input. The five classic sense-doors will quite clearly shut down in deep meditative states. Just because your eyes are closed does not make you blind. Light sensitivity remains, and as a side bonus of an optic nerve injury from a few years ago, I can ``see'' something (the damaged spot of the nerve) even when my eyes are shut and I'm in complete darkness. This blob which I ``see'' disappears in deep meditative states. Similarly, these deep states of meditation will eliminate audio input as well, which is easier for anyone to detect.

\begin{center}
  \LARGE{Vipassana: The Sledgehammer}
\end{center}

During the Day 4 discourse of a 10-day course, the imagery of the bridge is extended by what is on the far side of that bridge: a giant wall. The wall represents the remaining barrier between the conscious and unconscious mind -- and Vipassana is the tool used to break this barrier. By systematically examining the nervous system, outside-in, one digs through all the layers of unconscious habit as it manifests in bodily sensation.

\begin{figure}[h!]
  \centering
  \begin{subfigure}[b]{0.48\linewidth}
    \includegraphics[width=\linewidth]{images/wall.jpg}
  \end{subfigure}
  \begin{subfigure}[b]{0.48\linewidth}
    \includegraphics[width=\linewidth]{images/bridge.png}
  \end{subfigure}
  \label{fig:wall}
\end{figure}

And what comes of this examination? Sensations in the body are exploded into view and what we normally experience as a narrow band of sensation from the roughly unpleasant to the roughly pleasant: intense sensations like abdominal pain and annoying sensations like uncontrollable itching are gross sensations we normally consider unpleasant; intense sensations like sexual contact and light sensations like touching fur are gross sensations we normally consider pleasant.

>>>>

\usepackage{enumitem}

@@

\begin{titlepage}
   \vspace*{\stretch{1.0}}
   \begin{center}
     \Huge\textbf{Vipassana for Hackers}\\
     \Huge{Paper Four: Mastering Spacetime}\\
     \vspace{5cm}
     \large\textit{Steven Deobald}\\
     \large\textit{Version 0.1 DRAFT}\\
     \large\textit\today\\
     \vspace{5cm}
     \large\textit{}\\
   \end{center}
   \vspace*{\stretch{2.0}}
\end{titlepage}

@@
